---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
    theme: spacelab
---

## Loading the required libraries
```{r}
library(ggplot2) # for plotting the charts
library(chron) #for dealing with date and time in the dataset
```


## Loading and preprocessing the data
Load the data using the `read.csv` standard function:
```{r}
a <- read.csv(unz("activity.zip","activity.csv"), stringsAsFactors = FALSE)
```
The data-frame **a** must now contains 17568 observations of 3 variables. Let's check it:
```{r}
str(a)
```
The **date** column is of `character` data type. As we will have to deal with it as a date, let's convert **date** to the `Date` data type, then check the results (we also make certain the TZ is set to something universal so we have no timezone issues): 
```{r}
Sys.setenv(TZ='UTC')
a$date <- as.Date(a$date,"%Y-%m-%d")
str(a)
```
Let's check the interval variable by looking at the 50 first value:
```{r}
head(a$interval, 50)
```
What we see here is that the scale is not continuous. We go from 55 to 100 for a 5 minute interval. We need to adjust that so the scale equivalent to a time scale. By looking at the pattern, we see that the last two digit are the minutes from 0 to 55 and that the number for hundreds and probably thousands are the hours as they increment by one after the 55 minutes value. Lets see what appends for a whole day:
```{r}
ninter <- 24 * 60 / 5 #24h of 60 minutes devided by 5 to get the number of intervals for 1 day
head(a$interval, ninter + 1) #+1 to see what happen after midnight
```
This confirm our supposition. Let's convert that to a real proportional time interval by converting it into a `times` object of the *chron* library:
```{r}
a$interval <- times(((a$interval %% 100) /60 ) + #Minutes in fraction of hours
                     (a$interval %/% 100) # Hours
                    ) / 24 #Fraction of a day
```
It is better now. We have 3 variables:

*  **steps** : An integer value with the number of steps recorded in a 5 minutes interval.
*  **date** : The date on which the measurement was taken.
*  **interval** : A *time* object representing time after midnight when the measurement was taken.

## What is mean total number of steps taken per day?
We will now calculate the mean total number of steps per day. There is two steps to achieve this calculation:

1. Calculate the total number of steps for each day
2. Calculate the mean and median of these totals

Let's start by calculating the total number of steps for each day and have a look at the resulting data-frame :
```{r}
ts <- aggregate(steps ~ date, data = a,  sum)
str(ts)
```
Before calculating the mean and median, let's have a look of how the total number of steps per day are distributed:  
```{r NumberOfStepsPerDayHistogram}
ggplot() + geom_histogram(data = ts,aes(steps), bins = 11, fill = "red", color = "red", alpha = 0.5) +
           xlab("Number of steps per day") + ylab("Number of days")
```
  
The majority of the the total numbers of steps recorded for a given day seems to be around 10000. Let calculate the mean and the median to confirm that:
```{r}
mean(ts$steps)
median(ts$step)
```

## What is the average daily activity pattern?
Let's have a look of how the steps are distributed during the day by calculating the average number of steps per 5-minutes intervals and displaying it on a chart (as the date part is irrelevant we use an arbitrary date as our initial midnight):
```{r DailyActivityPattern}
ms <- aggregate(steps ~ interval, data = a, mean)
ggplot(data = ms) + aes(x = interval,
                        y = steps) +
                    geom_line() +
                    scale_x_chron(format = "%H:%M") +
                    xlab("Interval") +
                    ylab("Number of steps")
```

We see a spike in the chart. Let's find it's parameters (by using `()` around an assignment `<-` statement, R will also print the assigned value saving us to explicitly print the variable by retyping its name. This is particularly useful when you need to use a variable so you can put value in your text as we do below with the interval and number of steps):
```{r}
(maxms <- ms[which.max(ms$steps),])
```
So the average maximum number of steps occurs at the interval **`r maxms[1]`** with **`r round(maxms[2])`** steps on average for that time interval.

## Imputing missing values
There are some missing values in our data-set. This may have an impact on how we should interpret the data. So let's check them. How many NA's do we have:
```{r}
sapply(a, function(x) sum(is.na(x)))
```
So all the NA's are in the steps variable. What is the proportion of missing values in the step variable (as Boolean value are treated as 0 and 1 in calculation, the mean function will give us the proportion of NA's):
```{r}
(mis.step <- mean(is.na(a$step)))
```
So around `r round(mis.step * 100)`% of the value are missing. How can we fill in these missing value while minimizing the impact on the rest of the data. 
Let's find out a little more about these NA's. Are they spread in all the data? Are whole days missing ? Are specific intervals being more impacted than others?
Let's calculate the proportion of NA's for each day and have a look of what we have:
```{r}
na.d <- aggregate(is.na(steps) ~ date, data = a, mean)
summary(na.d)
```
So the minimum (Min.) is 0, meaning there are days without any missing values and the maximum (Max.) is 1 meaning that some days are full of NA's. Let's count the number of days that are neither full of NA's (1) nor without any NA's (0):
```{r}
nrow(subset(na.d, `is.na(steps)` != 0 & `is.na(steps)` != 1 ))
```
Good news, there are no partial data for any dates. Meaning that only full days of data are missing. How many days of data are missing:
```{r}
(misday <- sum(na.d$`is.na(steps)`))
```
Lets replace these `r misday` missing days of data by the mean of the other days for the corresponding interval. So basically we will assume that the days were there is no data available, the subject as performed an average day were each the number of steps for each interval is the mean of the days were data are available. We have already calculated the mean value per interval in the data frame **ms** but we will have to round the value to an integer to match the type of the other steps values:
```{r}
af <- merge(a, ms, by = "interval") # Add a column with the mean for the corresponding interval
af$steps.x <- ifelse(is.na(af$steps.x),as.integer(round(af$steps.y)), # replace NA's by the mean
                                       af$steps.x
                     )
af <- af[,c(2,3,1)] # drop the mean columns and reoder as in the original dataframe
colnames(af)[1] <- "steps" #rename the columan as in the original dataframe
```
Let's check if we still have missing values:
```{r}
sum(is.na(af$steps))
```
Good. Now let's compare the total steps per days before and after filling in the missing value
```{r CompareBeforeAndAfterHistogram}
tsf <- aggregate(steps ~ date, data = af,  sum)
ggplot() + geom_histogram(data = ts ,aes(steps), bins = 11, color = "red", alpha = 0.5, fill = "red") +
           geom_histogram(data = tsf,aes(steps), bins = 11, color = "green", alpha = 0.5) +
           xlab("Number of steps per day") + ylab("Number of days")
```

By overlaying the two histograms (original in red and, the one where NA has been filled with border in green and fill in gray), we can see that the number of days where the number of steps are all identical but for one bin. The central bin is now higher. 
Let's compare the means and medians:
```{r}
matrix(c(mean(ts$steps),   mean(tsf$steps),
         median(ts$steps), median(tsf$steps)),
       nrow = 2, 
       dimnames = list(c("Original (ts)", "Na Replaced (tsf)") ,
                       c("mean","median")
                       )
       )
```

The means and medians are practically identical. So filling in the missing value with the same interval means had nearly no impact on the overall mean or median for the days. This seems logical, the days we added where *mean* day so they impacted the number of *mean* days. There are `r misday` more in the central bin which correspond to the days where data was missing and that have been filled with data from a *average* day.

## Are there differences in activity patterns between weekdays and weekends?
First we need to add a variable to differentiate week days from weekend days then calculate the average number of steps per interval for the week days and weekend days separately:
```{r}
af$wd <- as.factor(ifelse(weekdays(af$date) %in% c("Saturday","Sunday"),"weekend","weekday"))
mswd <- aggregate(steps ~ interval + wd, data = af, mean)
```

Finally we compare the average number of steps during the day for the week days and weekend days :

```{r ActivityPatternWeekdaysAndWeekendComparison}
ggplot(data = mswd) + aes(x = interval,
                          y = steps) +
                      geom_line() +
                      scale_x_chron(format = "%H:%M") +
                      xlab("Interval") +
                      ylab("Number of steps") +
                      facet_grid("wd ~ .")
```

Looks like we are walking less in the beginning of a day during weekend than during the rest of the week and more in the afternoon.
